The importance of a complete statistical description of dependence between variables for ecology and related fields

### 2019-03-26 Paper.Rmd
asym sens results explained

The second mechanism we explored, which was also summarized in the 
Introduction,
relates to Liebig's law of the minimum, or, more generally, to the fact that 
environmental variables can often influence ecological variables through nonlinear 
relationships. If an environmental variable influences an ecological variable
more strongly in one of its tails, we explored using a model 
whether the ecological variable
could then exhibit asymmetric tail dependence between measurements made in two
locations. Let $E_i(t)$ again be an ecological variable measured in locations
$i=1,2$. Assume $E_i(t)$ has dynamics that follow 
$E_i(t+1)=bE_i(t)+g(\epsilon_i(t))+a \delta_i(t)$, where $g$ is a function
describing how the environmental variable $\epsilon_i(t)$ influences
$E_i(t)$, and the $\delta_i(t)$ are standard-normally distributed 
local noise, independent across time and sampling locations. We used $a=0.2$, and 
$b=0.1, -0.1, 0.9, -0.9$ in different simulations. 
Negative values of $b$ correspond to overcompensating dynamics 
and positive values correspond to undercompensating dynamics; larger-magnitude $b$
mean slower return time of the model to equilibrium after a perturbation. 
The $(\epsilon_1(t),\epsilon_2(t))$ were drawn, independently through
time, from a bivariate normal distribution with $\var(\epsilon_i)=1$ and 
$\cov(\epsilon_1,\epsilon_2)=0.8$. 
We used $g$ equal to $g_1$ or $g_2$ in two different simulations:
$g_1(\epsilon)$ equals $\epsilon$ if $\epsilon<0$ and $0$ otherwise; and 
$g_2(\epsilon)$ equals $\epsilon$ if $\epsilon>0$ and $0$ otherwise. Thus
$g_1$ represent environmental effects that occur only for negative 
values of the environmental variables, and $g_2$ represent environmental effects 
that occur only for positive values. Note that both $(\epsilon_1,\epsilon_2)$
and $(\delta_1,\delta_2)$ have N copula structure, being drawn from 
bivariate normal distributions. Hence the Moran-like mechanism discussed
previously cannot be operating in this model.

For each value of $b$, we simulated the model once using $g_1$ and once using $g_2$, for 5000 time
steps in both cases, retaining $E_1(t)$ and $E_2(t)$ for the final 500 time 
steps of each simulation. We applied our model selection algorithms and 
non-parametric statistics, which were previously applied to 
data to answer Q1, to the 
simulation outputs to discover if the model could
give rise to non-N copula structure and asymmetric tail dependence in the 
ecological variable, in spite of the absence of such structure in the driver
environmental variables.

@@

For strong noise and using our nonlinear model (methods for Q2), 
copula statistics, generally speaking, were approximately similar 
between noise and model
outputs; however, similarity was reduced relative to the AR(1) 
and weak-noise cases, and, for a few simulations, tail dependence 
asymmetry statistics
even had opposite signs for noise and model outputs 
(Figs \ref{SM-fig_Cause4copula_C_kend_sRicker_nsd_1}-\ref{SM-fig_Cause4copula_SJ_kend_sRicker_nsd_1}). For instance, using a C copula with a large Kendall $\tau$,
$\cor_l - \cor_u$ was slightly positive for noise, but slightly negative for 
population model outputs (Fig. \ref{SM-fig_Cause4copula_C_kend_sRicker_nsd_1}); 
and likewise for SG and SJ copulas (Figs \ref{SM-fig_Cause4copula_SG_kend_sRicker_nsd_1}, 
ref{SM-fig_Cause4copula_SJ_kend_sRicker_nsd_1}). Dynamics were
strongly nonlinear for the model parameters we used, and nevertheless similarities
between noise and model-output copula structure, according to the statistics 
we used, were more noticable than discrepancies.
We again repeated the analysis with $r=1.3$. Discrepancies between noise and population
copula structure in this case were often glaring. Apparently noise of standard deviation 1
interacted very strongly with model nonlinearities when the model was in its overcompensatory 
regime.

Our model with environmental effects which differed in the
two tails could produce simulated ecological variables with visually apparent asymmetry 
of tail dependence and non-N copula structure, depending on the value of $b$ used. 
For $b=0.1$ (Fig. \ref{fig_ms_asym_b_0.1}) and $b=-0.1$ (Fig. \ref{SM-fig_ms_asym_b_-0.1}), 
for both $g_1$ and $g_2$,
asymmetric tail dependence and non-N copula structure were visually apparent.
Lower-tail (respectively, upper-tail) effects of the noise ($g_1$, resp., $g_2$) 
produced lower-tail (resp., upper-tail) dependence in the ecological variable.
Model selection results (Table \ref{SM-tab_summary_stat_asym_sens}) and
our nonparametric statistics (Tables \ref{SM-tab_stat_npa_asym}) 
statistically confirmed the visual result for $b=0.1, -0.1$. Results were more
complex for $b=0.9$ or $-0.9$, indicating that strongly autocorrelated (or strongly 
negatively autocorrelated) dynamics interact with the effect we have described.
Ecological-variable copulas were strongly non-N for $g_1$, but not for $g_2$ 
and asymmetries of tail dependence were not particularly noticable 
(Figs \ref{SM-fig_ms_asym_b_0.9}-\ref{SM-fig_ms_asym_b_-0.9} and Tables 
\ref{SM-tab_summary_stat_asym_sens}-\ref{SM-tab_stat_npa_asym}).
<!--***DAN: Shya, this has got to be an error. Results should be symmetric
with respect to the choice of g_1 and g_2 because the model is symmetric. Let's discuss.
I think there is likely a bug affecting Figs S51 and 51, and Tables S47-48, and 
possibly all results based on the same model.-->

>>>>

The second mechanism we explored, which was also summarized in the 
Introduction,
relates to Liebig's law of the minimum, or, more generally, to the fact that 
environmental variables can often influence ecological variables through nonlinear 
relationships. If an environmental variable influences an ecological variable
more strongly in one of its tails, we explored using a model 
whether the ecological variable
could then exhibit asymmetric tail dependence between measurements made in two
locations. Let $E_i(t)$ again be an ecological variable measured in locations
$i=1,2$. Assume $E_i(t)$ has dynamics that follow 
$E_i(t+1)=bE_i(t)+g(\epsilon_i(t))+a \delta_i(t)$, where $g$ is a function
describing how the environmental variable $\epsilon_i(t)$ influences
$E_i(t)$, and the $\delta_i(t)$ are standard-normally distributed 
local noise, independent across time and sampling locations. We used $a=0.2$, and 
$b=0.1, -0.1, 0.5, -0.5$ in different simulations. 
Negative values of $b$ correspond to overcompensating dynamics 
and positive values correspond to undercompensating dynamics; larger-magnitude $b$
mean slower return time of the model to equilibrium after a perturbation. 
The $(\epsilon_1(t),\epsilon_2(t))$ were drawn, independently through
time, from a bivariate normal distribution with $\var(\epsilon_i)=1$ and 
$\cov(\epsilon_1,\epsilon_2)=0.8$. 
We used $g$ equal to $g_1$ or $g_2$ in two different simulations:
$g_1(\epsilon)$ equals $\epsilon$ if $\epsilon<0$ and $0$ otherwise; and 
$g_2(\epsilon)$ equals $\epsilon$ if $\epsilon>0$ and $0$ otherwise. Thus
$g_1$ represent environmental effects that occur only for negative 
values of the environmental variables, and $g_2$ represent environmental effects 
that occur only for positive values. Note that both $(\epsilon_1,\epsilon_2)$
and $(\delta_1,\delta_2)$ have N copula structure, being drawn from 
bivariate normal distributions. Hence the Moran-like mechanism discussed
previously cannot be operating in this model.

For each value of $b$, we simulated the model once using $g_1$ and once using $g_2$, for $25000$ time
steps in both cases, retaining $E_1(t)$ and $E_2(t)$ for the final $2500$ time 
steps of each simulation. We applied our model selection algorithms and 
non-parametric statistics, which were previously applied to 
data to answer Q1, to the 
simulation outputs to discover if the model could
give rise to non-N copula structure and asymmetric tail dependence in the 
ecological variable, in spite of the absence of such structure in the driver
environmental variables.

@@

For strong noise and using our nonlinear model (methods for Q2), 
copula statistics, generally speaking, were approximately similar 
between noise and model
outputs; however, similarity was reduced relative to the AR(1) 
and weak-noise cases, and, for a few simulations, tail dependence 
asymmetry statistics
even had opposite signs for noise and model outputs 
(Figs \ref{SM-fig_Cause4copula_C_kend_sRicker_nsd_1}-\ref{SM-fig_Cause4copula_SJ_kend_sRicker_nsd_1}). For instance, using a C copula with a large Kendall $\tau$,
$\cor_l - \cor_u$ was slightly positive for noise, but slightly negative for 
population model outputs (Fig. \ref{SM-fig_Cause4copula_C_kend_sRicker_nsd_1}); 
and likewise for SG and SJ copulas (Figs \ref{SM-fig_Cause4copula_SG_kend_sRicker_nsd_1}, 
\ref{SM-fig_Cause4copula_SJ_kend_sRicker_nsd_1}). Dynamics were
strongly nonlinear for the model parameters we used, and nevertheless similarities
between noise and model-output copula structure, according to the statistics 
we used, were more noticable than discrepancies.
We again repeated the analysis with $r=1.3$. Discrepancies between noise and population
copula structure in this case were often glaring. Apparently noise of standard deviation 1
interacted very strongly with model nonlinearities when the model was in its overcompensatory 
regime.

Our model with environmental effects which differed in the
two tails could produce simulated ecological variables with visually apparent asymmetry 
of tail dependence and non-N copula structure, depending on the value of $b$ used. 
For $b=0.1$ (Fig. \ref{fig_ms_asym_b_0.1}) and $b=-0.1$ (Fig. \ref{SM-fig_ms_asym_b_-0.1}), 
for both $g_1$ and $g_2$,
asymmetric tail dependence and non-N copula structure were visually apparent.
We also got similar results from simulations with $b=\pm 0.5$ (Figs \ref{SM-fig_ms_asym_b_0.5} - 
\ref{SM-fig_ms_asym_b_-0.5}).
Lower-tail (respectively, upper-tail) effects of the noise ($g_1$, resp., $g_2$) 
produced lower-tail (resp., upper-tail) dependence in the ecological variable.
Model selection results (Table \ref{SM-tab_summary_stat_asym_sens}) confirmed that
non-Normal copula was always best fit for $b= \pm 0.1$ and $\pm 0.5$, but their model 
AIC-wieght averaged tail dependence could not reflect the asymmetry for each case 
when that dependnece was at too much extreme end.
However, our nonparametric statistics (Tables \ref{SM-tab_stat_npa_asym}) focused on those extreme tips
and overcome the shortcomings of model selection approach which fit data based on maximum likelihood.

<!--***DAN: Shya, this has got to be an error. Results should be symmetric
with respect to the choice of g_1 and g_2 because the model is symmetric. Let's discuss.
I think there is likely a bug affecting Figs S51 and 51, and Tables S47-48, and 
possibly all results based on the same model.-->
